---
title: "What is Due & When?"
author: "Alison Hill"
subtitle: Summer 2018
---

You are expected to come to each class meeting prepared, which means that you will have done the readings and the DataCamp assignments that are due for that date *before* coming to class. 

## Readings

Readings will primarily be from two free, open-source, completely online textbooks:

1. [**MODERN DIVE into Data with R (MD)**](http://moderndive.netlify.com) 
1. [**R for Data Science (R4DS)**](http://r4ds.had.co.nz) 

These books will be supplemented by blog posts, YouTube videos, and several open-access articles, including from:

- A special PEERJ issue called ["Practical Data Science for Stats"](https://peerj.com/collections/50-practicaldatascistats/), and
- [Nature: Points of Significance](https://www.nature.com/collections/qghhqm/pointsofsignificance)

## DataCamp

Be sure to check the **DataCamp** link above for more details on the DataCamp (DC) assignments.  More details on the Problem Sets (PS) are available in the link above for Problem Sets.

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(DT)
due <- read_csv("due.csv") %>%
  mutate(Date = paste(Month, Day)) %>%
  mutate(`DataCamp Due` = ifelse(material == "-" | grepl("Quiz", material) | is.na(material_link),
    material,
    paste0("<a href='", material_link, "'>", material, "</a>"))) %>% 
  mutate(`Problem Set Due` = ifelse(ps == "-" | is.na(ps) | is.na(ps_link),
    ps,
    paste0("<a href='", ps_link, "'>", ps, "</a>"))) %>% 
  select(Weekday, Date, `CM / Block`, `Readings Due` = Reading, `DataCamp Due`, `Problem Set Due`)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(knitr:::is_html_output()){
  due %>% 
    datatable(rownames = FALSE,
              class = 'cell-border stripe',
              filter = list(position = 'top'),
              options = list(pageLength = nrow(due), 
                             autoWidth = TRUE,
                             bInfo = FALSE,
                             paging = FALSE),
              escape = FALSE)
} else {
  knitr::kable(due, booktabs = TRUE, longtable = TRUE)
}
```