---
title: "Class Meeting 1.3 (cm013)"
subtitle: "`babynames` wrangling drills"
author: "Alison Hill"
date: "6/28/2018"
output:
  html_document:
    code_folding: hide
    theme: flatly
    highlight: pygments
    toc: true
    toc_float: true
---


```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set(error = TRUE, comment = NA, warnings = FALSE, messages = FALSE, tidy = FALSE, cache = TRUE)
```

```{r load_packages, include = F}
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(ggplot2)))
suppressWarnings(suppressMessages(library(babynames)))
suppressWarnings(suppressMessages(library(psych)))
suppressWarnings(suppressMessages(library(tidyr)))
```


1. Create a new dataframe that only includes years since 2000.
```{r}
newest <- babynames %>%
  filter(year >= 2000)
```

2. What name has been used for the most number of years (when used for a single gender)? 

_Hint 1:_ find the total number of years each name/sex combination has been used.
_Hint 2:_ what is the maximum total number of years any name has been used?

```{r}
years <- babynames %>%
  group_by(name, sex) %>%
  summarise(use_years = n()) %>% 
  arrange(desc(use_years))
```
It is a 934 way tie

3. How many names have only been used one year? 
```{r}
years %>%
  ungroup %>%
  group_by(sex) %>%
  filter(use_years == 1)
```
There are 20,472 of them.

4. Create a new column that displays the number of genders the name was used for each year. Note: not recommended for slower computers.

```{r}
boygirl <- babynames %>%
  group_by(year, name) %>%
  mutate(both = n())
```

5. Make a data set of names that have been both boy and girl names. 

_Hint:_ use `distinct()`

```{r}
new <- boygirl %>%
  filter(both == 2) %>%
  distinct(name)
```

6. For each year, display the total number of names that were used. Treat boy and girl versions of the same name as two separate names.

```{r}
new <- babynames %>%
  group_by(year, sex) %>%
  summarise(num_names = n())
```
about 2000 names were recorded for each year

7. Which name received the largest percentage of any name for any year (consider boy and girl names as distinct) 

```{r}
babynames %>%
  filter(prop == max(prop))
```
John in 1880

8. Which girl's name received the largest percentage of any girl's name for any year? 

```{r}
babynames %>%
  filter(sex == "F") %>%
  filter(prop == max(prop))
```
Mary in 1880

9. Display the average percentage each name received during all the years it was used. Treat girl and boy versions of the same name as different names. 

```{r}
prop <- babynames %>%
  group_by(name, sex) %>%
  summarise(avg_prop = mean(prop))
```

10. Which name recorded in the data set has been out of use for the longest time? 

```{r}
last <- babynames %>%
  group_by(name) %>%
  summarise(last_use = max(year)) 

last %>%
  filter(last_use == min(last_use))
```
2 names have been out of use since 1881

11. In a new column, display the total number of years a name has been used (as either a boys name or a girls name). 

```{r}
new <- babynames %>%
  group_by(name) %>%
  mutate(num_years = length(year))
```

12. In general, are names that have been used for both boys and girls more popular for boys or girls? 

```{r eval = F}
bg_names <- prop %>%
  filter(name %in% both) %>%
  group_by(name) %>%
  summarise(diff = avg_prop[sex=="M"] - avg_prop[sex =="F"])
mean(bg_names$diff >= 0)
```
More popular for girls (new$diff $gt; 0 is TRUE (i.e, 1) when the boys version is more popular. This occurred less than half the time (i.e. mean(new$diff $gt; 0) < 0.5)

13. What name has earned the most percentage points in a year for any name since 2000? 
```{r}
newest %>%
  filter(prop == max(prop))
```
Jacob in 2000
